Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-02-19 RNARedPrint-Bioinfo2018.tex
More about pseudoknots; optimization run interpretation

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
sequences of length $n$. An \Def{RNA (secondary) structure $R$ of
  length $n$} is a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1''). We call an RNA
structure $R$ \Def{non-crossing}, iff it does not contain any two
different base pairs $(i,j)$ and $(i',j')$ such that
$i\leq i'\leq j \leq j'.$ \Def{Valid base pair} must pair bases from
the $\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
As will be explicit in subsequent sections, our approach relies on breaking down the energy function $E_\R(S)$ into additive components, each depending on only few sequence positions. Given $\R$, we express $E_\R(S)$
as the sum of energy contributions $f(S)$ over a set $\F$ (of
functions $f:\S_n\to\real$), s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This
captures realistic RNA energy models, while bounding the
dependencies to sequence positions introduced by each single $f$. Formally, define the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g, we assume that $T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
root $r$ connected to a node in each connected component of $T$.  Now,
we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
$r$; define $T_r(u)$ as the induced subtree of
$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
messages along these directed edges $u\to v$ (i.e. always from some
child $u$ to its parent $v$). Each message is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure
$\sample(u,\val)$, which returns a partial sequence that determines
the positions of the subtree rooted at $u$ from the
Boltzmann-distribution.  Called on $r$ and the empty partial sequence,
the procedure returns a Boltzmann-distributed random sequence.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly at random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of the quality of Boltzmann designs over uniform ones, suggesting a superior potential for optimization ($\sim 32\%$ avg improvement, Table~\ref{tab:benchmark-results}, optimized). This improvement is however more modest than in the absence of optimization, but we attribute this relative decay to the uniform nature of the resampling strategy, which somehow ends up canceling the benefits of the initial Boltzmann distribution, by uniformizing the sequence over longer runs. This suggests an alternative resampling strategy based on Boltzmann-distributed steps.

>>>>

An \Def{RNA sequence $S$} is a word over the \Def{nucleotides
  alphabet} $\Sigma=\{\Ab,\Cb,\Gb,\Ub\}$; let $\S_n$ denote the set of
sequences of length $n$. An \Def{RNA (secondary) structure $R$ of
  length $n$} is a set of \Def{base pairs} $(i,j)$, where
$1\leq i<j\leq n$, where for all different $(i,j), (i',j')\in R$:
$\{i,j\}\cap\{i',j'\}=\emptyset$ (``degree $\leq$ 1'').
% 
% We call an RNA structure $R$ \Def{non-crossing}, iff it does not
% contain any two different base pairs $(i,j)$ and $(i',j')$ such that
% $i\leq i'\leq j \leq j'.$
% 
\Def{Valid base pair} must pair bases from the
$\B:=\left\{\{\Ab,\Ub\},\{\Gb,\Cb\},\{\Gb,\Ub\}\right\}.$
Consequently, $S$ is \Def{valid} for $R$, iff $\{S_i,S_j\}\in \B$ for
all $(i,j)\in R$.

@@

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
  As will be explicit in subsequent sections, our approach relies on
  breaking down the energy function $E_\R(S)$ into additive
  components, each depending on only few sequence positions. Given
  $\R$, we express $E_\R(S)$ as the sum of energy contributions $f(S)$
  over a set $\F$ (of functions $f:\S_n\to\real$),
  s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This captures realistic RNA
  energy models---including nearest neighbor models, e.g.~\citet{Turner2009}, and even
  pseudoknot models, e.g. \citet{Andronescu2010}, while bounding the dependencies to
  sequence positions introduced by each single $f$. Formally, define
  the \Def{dependencies} $\dep(f)$ of $f$ as a minimal set of sequence
  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
  sequences $S$ and $S'$ agreeing at all positions in $\I$.

@@

Let us now consider the \Def{computation of the partition function}.
Given is the RNA design network $\network=(\X,\F)$ and its cluster
tree decomposition $(T,\chi,\phi)$.  W.l.o.g., we assume that $T$ is
connected and contains a dedicated node $r$, with
$\chi(r)=\varnothing$ and $\phi(r)=\varnothing$, added as a virtual
root $r$ connected to a node in each connected component of $T$.  Now,
we consider the set of directed edges $\edgesToR{}$ of $T$ oriented to
$r$; define $T_r(u)$ as the induced subtree of
$u$. Algorithm~\ref{alg:pf} computes the partition function by passing
messages along these directed edges $u\to v$ (i.e. always from some
child $u$ to its parent $v$). Each message is a function that depends
on the positions $\dep(m)\subseteq \X$ and yields a partition function
in $\real\cup\{\infty\}$. The message from $u$ to $v$ represents the
partition functions of the subtree of $u$ for all possible partial
sequences in $\partseqs(\separator{u}{v})$. Induction over $T$ lets us show
the correctness of the algorithm
(Supp. Mat.~\ref{appsec:correctness}).  After running
Alg.~\ref{alg:pf}, multiplying the 0-ary messages sent to the root $r$
yields the total partition function:
\begin{math}
  \partfun{\network} = \prod_{(u\to{}r)\in T} \evalfor{\Message{u}{r}}{\varnothing}.
\end{math}

@@

The partition functions can then direct a \Def{stochastic backtrack} to achieve \Def{Boltzmann sampling of sequences}, such that one samples from the Boltzmann distribution of a given design network $\network$. The sampling algorithm assumes that the cluster tree was expanded and the messages $\Message{u}{v}$ for the edges in $\edgesToR{}$ are already generated by Algorithm \ref{alg:pf} for the expanded cluster tree.
%
Algorithm~\ref{alg:sampling} defines the recursive procedure
$\sample(u,\val)$, which returns---randomly drawn from the
Boltzmann distribution---a partial sequence that determines all
sequence positions in the subtree rooted at $u$.  Called on $r$ and
the empty partial sequence, which does not determine any positions,
the procedure samples a random sequence from the Boltzmann
distribution.

@@

Moreover, for each generated seed sequence, we optimized the objective function $(\ref{eq:blueprintobjective})$ through an adaptive  greedy walk consisting of 500 steps. At each step, we resampled (uniformly random) the positions of a randomly selected component in the base pair dependency graph, and accepted the modification only if it resulted in a gain, as described in \Software{RNAblueprint}~\citep{Hammer2017}.
%
Once again, for all instances, we observe a positive improvement of
the quality of Boltzmann designs over uniform ones, suggesting a
superior potential for optimization ($\sim 32\%$ avg improvement,
Table~\ref{tab:benchmark-results}, optimized). Notably, our subsequent
optimization runs, which are directly inherited from the uniformly
sampling RNABlueprint software, partially level the advantages of
Boltzmann sampling. In future work, we hope to improve this aspect by
exploiting Boltzmann sampling even during the optimization run.
